Honey Bee Colonies Impacted by Varroa, American Foulbrood and Global Warming

### 2024-08-03 starter-analysis-exercise/products/report/manuscript_final-2.qmd
Update format

### Composition of Honey

@@

### Spread of Varroa Mites

@@

The map reveals the progressive spread of Varroa mites from the
mid-1980s to the mid-1990s, affecting bee colonies across the country.
The earliest appearances were concentrated in specific regions and
gradually spread to more states over time. This visual representation
highlights the increasing geographical distribution of Varroa mites and
underscores the growing challenge they pose to beekeeping and
agricultural industries in the United States.

[@varroamap] ***Varroa Map of infestation periods***

@@

@@

Bee colonies maintained by beekeepers are considered livestock by the
USDA due to their ability to produce honey, a consumable food item, and
their essential role in assisting farmers with pollination crop seasons.
Given the importance of bee colonies in agriculture, it was logical to
source data from the following two authoritative websites: 1. USDA
National Agricultural Statistics Service (NASS): This site provides
comprehensive agricultural data, including statistics on honey
production and colony health. 2. Bee Informed Partnership: This site
offers detailed insights and research on bee colony management and
health, contributing valuable information on the status and trends of
bee populations. [@usda_honey_bees]

Index Catalog // USDA Economics, Statistics and Market Information
System. (n.d.-a).
https://usda.library.cornell.edu/catalog?f%5Bkeywords_sim%5D%5B%5D=honey+bees&locale=en\
USDA - National Agricultural Statistics Service - Surveys - honey bee
surveys and reports. (n.d.).
https://www.nass.usda.gov/Surveys/Guide_to_NASS_Surveys/Bee_and_Honey/

## Questions/Hypotheses to be addressed

Hypothesis: "The negative impacts of mites, bacterium, and global
warming have detrimental effects on honeybee colonies in the United
States and Texas, which in turn will lead to a decline in honey
production and negatively impact food production." This hypothesis can
be tested and validated through a visualization of outcomes using R,
demonstrating the relationship between these factors and their effects
on honeybee colonies.

@@

![](/images/dead-bees-hive-colony-collapse-600nw-2289939105.webp)
{{< pagebreak >}}

# Methods

*Describe your methods. That should describe the data, the cleaning
processes, and the analysis approaches. You might want to provide a
shorter description here and all the details in the supplement.*

@@

# Create the flowchart using DiagrammeR
grViz("
  digraph Data_Analysis_Flowchart {
    graph [layout = dot, rankdir = LR, label = 'Workflow Schematic', fontsize = 45, labelloc = 't']

    node [style = filled, fontsize = 30, width = 2, height = 2]

@@

Data was derived from the United States Department of Agriculture (USDA) website, which permits data sets to be downloaded in CSV(comma-separated values) format.

We decided to clean out our data from a few different datasets. We had
to remove blank spaces and columns that were not pertinent to our
analysis. We then filtered out other observations that did not directly
deal with the data we are exploring. We are looking for cause of death
to bee colonies and how they are affected by mites and climate change so
we wanted to single out data that represented the losses so we can
explore the different states by year and determine how the colonies were
affected.

@@

### Scatterplot of Value by State ANSI

The fourth image is a scatterplot showing values for different states
(ANSI coded) across the years 2015 to 2017. Key points include: - Each
color represents a different year (2015: Red, 2016: Green, 2017:
Blue). - The scatterplot reveals the distribution of values across
states and highlights any trends or patterns over the three years.

This scatterplot indicates that while some states experience
consistently high or low values, others show significant variability
year over year. This could suggest regional differences in factors
affecting bee colony health.

@@

### Boxplot of Value by Period and Data Item

The fifth image is a boxplot comparing the values for "LOSS, COLONY
COLLAPSE DISORDER" and "LOSS, DEADOUT" across different periods of the
year: - **APR THRU JUN** - **JAN THRU MAR** - **JUL THRU SEP** - **OCT
THRU DEC**

Key insights include: - Seasonal trends in colony losses, with some
periods showing higher median losses or more variability. - Outliers
indicate extreme loss events occurring in specific periods.

This seasonal analysis suggests that certain times of the year are more
prone to higher losses, potentially due to environmental factors,
resource availability, or management practices.

@@

##The full analysis is still pending###

@@

# References

>>>>

# Plot with ggplot2
ggplot(honey_data, aes(x = "", y = percentage, fill = component)) +
  geom_bar(width = 0.8, stat = "identity", color = "white") +
  coord_polar("y", start = 0) +
  scale_fill_manual(values = honey_data$color) +
  geom_text(aes(x=1.8, label = paste0(percentage, "%")), position = position_stack(vjust = 0.6), size = 3, color = "black") +
  labs(
    title = "COMPOSITION OF HONEY",
    subtitle = "Illustration of honey components by percentage",
    fill = NULL
  ) +
  theme_void() +
  theme(
    plot.title = element_text(size = 22, face = "bold", hjust = 0.5, family = "lobster"),
    plot.subtitle = element_text(size = 14, hjust = 0.5),
    legend.position = "bottom",
    legend.text = element_text(size = 12)
  )
```
### Composition of Honey

@@

```
### Spread of Varroa Mites

@@

The map reveals the progressive spread of Varroa mites from the
mid-1980s to the mid-1990s, affecting bee colonies across the country.
The earliest appearances were concentrated in specific regions and
gradually spread to more states over time. This visual representation
highlights the increasing geographical distribution of Varroa mites and
underscores the growing challenge they pose to beekeeping and
agricultural industries in the United States.
***[@varroamap] Varroa Map of infestation periods***

@@

```
The chart provided illustrates the changes in temperature anomalies over the years, from 1900 to approximately 2025. The temperature anomaly, measured in degrees Celsius (°C), is plotted on the y-axis, while the years are plotted on the x-axis. The data is divided into three different measurements: Earth's surface, lower troposphere (RSS), and lower troposphere (UAH), each represented by different colors (blue, red, and green respectively).

### Detailed Description:
- **Trend Analysis**:
  - From 1900 to around 1975, the Earth's surface temperature shows considerable variability with some notable periods of cooling and warming.
  - Post-1975, there is a distinct upward trend in temperature anomalies for all three measurements, indicating a general warming trend.
- **Measurement Comparison**:
  - The Earth's surface (blue line) shows the most extended historical data and fluctuates more in the earlier years compared to the troposphere measurements.
  - The lower troposphere measurements (RSS in red and UAH in green) start around 1975 and generally follow the same upward trend as the surface measurements but with less variability.
  - The RSS and UAH measurements align closely with each other post-2000, indicating consistent trends in the lower troposphere temperature anomalies.
- **Anomalies & Peaks**:
  - The Earth's surface shows significant peaks and troughs, especially in the early 20th century and around the year 2000.
  - The lower troposphere measurements show peaks and align closely, particularly after the year 2000, reflecting similar patterns in temperature anomalies.

### Key Points:
- The overall trend shows an increase in temperature anomalies from 1900 to 2025.
- Significant variability is seen in Earth's surface temperatures in the early years.
- Post-1975, a clear warming trend is observed across all measurements.
- Lower troposphere data (RSS and UAH) begins around 1975 and follows the Earth's surface trend with less variability.
- Post-2000, the RSS and UAH data closely align, indicating consistent temperature trends in the lower troposphere.
- There are noticeable peaks in the temperature anomalies around the year 2000 and later, reflecting periods of significant temperature increases.

### Impact on Bees, Varroa Mites, and Colony Loss:
- **Bees**:
  - **Thermal Stress**: Increased temperature anomalies can cause thermal stress on bees, affecting their ability to forage, navigate, and perform essential colony tasks.
  - **Floral Resources**: Warming temperatures can alter the availability and distribution of floral resources, impacting bee nutrition and health.
  - **Reproduction Cycles**: Changes in temperature can disrupt the reproductive cycles of bees, potentially leading to mismatches with the availability of pollen and nectar.

- **Varroa Mites**:
  - **Increased Reproduction**: Warmer temperatures can accelerate the reproduction rates of varroa mites, exacerbating infestations within bee colonies.
  - **Extended Activity Period**: Higher temperatures can lengthen the active period of varroa mites throughout the year, leading to prolonged periods of stress on bee populations.

- **Colony Loss**:
  - **Health Decline**: Combined effects of thermal stress on bees and increased varroa mite infestations can lead to a decline in colony health.
  - **Colony Collapse Disorder (CCD)**: Temperature anomalies can contribute to conditions that favor Colony Collapse Disorder, where the majority of worker bees disappear, leaving behind the queen and immature bees.
  - **Mortality Rates**: Sustained temperature increases can lead to higher mortality rates within bee colonies, significantly impacting beekeeping and agricultural pollination services.

### Conclusion:
The chart highlights a clear trend of increasing temperature anomalies, which can have profound implications on bee populations, varroa mite dynamics, and overall colony health. Understanding these trends is crucial for developing strategies to mitigate the adverse effects on bees and ensure the sustainability of pollination services essential for ecosystems and agriculture.

@@

### Data Sources

Bee colonies maintained by beekeepers are classified as livestock by the USDA due to their production of honey, a consumable food item, and their critical role in pollination during crop seasons. Given the importance of bee colonies in agriculture, we sourced data from two authoritative websites:

1. **USDA National Agricultural Statistics Service (NASS)**:
   - Provides comprehensive agricultural data, including statistics on honey production and colony health.
   - Source: [USDA - National Agricultural Statistics Service - Surveys - honey bee surveys and reports](https://www.nass.usda.gov/Surveys/Guide_to_NASS_Surveys/Bee_and_Honey/).

2. **Bee Informed Partnership**:
   - Offers detailed insights and research on bee colony management and health, contributing valuable information on the status and trends of bee populations.
   - Source: [Index Catalog // USDA Economics, Statistics and Market Information System](https://usda.library.cornell.edu/catalog?f%5Bkeywords_sim%5D%5B%5D=honey+bees&locale=en).

These sources provided reliable and extensive data necessary for analyzing the factors affecting bee colonies, honey production, and the broader implications for agriculture.

## Questions/Hypotheses to be addressed
**Hypothesis**:
"The negative impacts of mites, bacterium, and global warming have detrimental effects on honeybee colonies in the United States and Texas, which in turn will lead to a decline in honey production and negatively impact food production." This hypothesis can be tested and validated through a visualization of outcomes using R, demonstrating the relationship between these factors and their effects on honeybee colonies.

**Null Hypothesis (H0)**:
The impacts of mites, bacterium, and global warming do not have a significant detrimental effect on honeybee colonies in the United States and Texas, and there is no consequent decline in honey production or negative impact on food production.

**Alternative Hypothesis (H1)**:
The impacts of mites, bacterium, and global warming have a significant detrimental effect on honeybee colonies in the United States and Texas, leading to a decline in honey production and negatively impacting food production.

These hypotheses will be tested using various data visualization techniques in R, allowing us to explore and validate the relationships between these factors and their effects on honeybee colonies.

@@

![](/images/dead-bees-hive-colony-collapse-600nw-2289939105.webp)

#Here are the BibTex entries for citations:

@Article{hivecollapse,
  author = {Brian Dennis and William P. Kemp},
  title = {How Hives Collapse: Allee Effects, Ecological Resilience, and the Honey Bee},
  journal = {PLOS ONE},
  year = {2016},
  volume = {11},
  number = {2},
  pages = {e0150055},
  doi = {10.1371/journal.pone.0150055},
  url = {https://doi.org/10.1371/journal.pone.0150055}
A study performed to determine various elements impacting bee hive collapse}

@USDAWebsite{varroamap,
  author = {sdns6mchl4},
  title = {Varroa mite spread in the United States},
  year = {2016},
  month = {February 24},
  url = {https://www.beesource.com/threads/varroa-mite-spread-in-the-united-states.365462/},
  note = {Beesource Beekeeping Forums}
Map highlighting states of Varroa infestation from periods 1986-1995}

@USDAWebsite{usda_honey_bees,
  author = {{USDA Economics, Statistics and Market Information System}},
  title = {Index Catalog},
  year = {n.d.-a},
  url = {https://usda.library.cornell.edu/catalog?f%5Bkeywords_sim%5D%5B%5D=honey+bees&locale=en},
  note = {Accessed: 2024-07-31}
Data sets}

@USDAWebsite{usda_nass,
  author = {{USDA - National Agricultural Statistics Service}},
  title = {Surveys - honey bee surveys and reports},
  year = {n.d.},
  url = {https://www.nass.usda.gov/Surveys/Guide_to_NASS_Surveys/Bee_and_Honey/},
  note = {Accessed: 2024-07-31}
Data sets}

@EPAWebsite{epa_climate,
  author = {{U.S. Environmental Protection Agency}},
  title = {Climate Change Indicators: U.S. and Global Temperature},
  year = {2024},
  howpublished = {\url{https://www.epa.gov/climate-indicators/climate-change-indicators-us-and-global-temperature}},
  note = {Accessed: 2024-08-01}
Climate change from periods of 1900-2024}

@article{2024warmer,
  title={Warmer autumns and winters could reduce honey bee overwintering survival with potential risks for pollination services},
  author={Rajagopalan, K. and DeGrandi-Hoffman, G. and Pruett, M. and others},
  journal={Scientific Reports},
  volume={14},
  pages={5410},
  year={2024},
  publisher={Nature Publishing Group},
  doi={10.1038/s41598-024-55327-8}
Peer reviewed article concering global warming possibly impacting bee hive survivability and viability during winter periods}

# Methodology

The United States Department of Agriculture (USDA) website served as our data source. We downloaded CSV files containing vital information on honey production, colony health, and factors impacting bee populations. By using USDA data, we ensured our analysis was based on reliable, comprehensive, and relevant records widely recognized in agricultural and ecological research.

### Data Cleaning and Preparation
We integrated data from multiple datasets to ensure a comprehensive analysis. The cleaning process involved several key steps:

1. **Removal of Irrelevant Data**:
   - **Blank Spaces**: We eliminated any blank spaces within the datasets to ensure data integrity.
   - **Non-Pertinent Columns**: Columns that were not directly related to our analysis objectives were removed to streamline the dataset.

2. **Filtering Observations**:
   - **Focus on Relevant Data**: We filtered out observations that did not directly pertain to our study, specifically those not related to the causes of bee colony losses.
   - **Loss Data**: The dataset was refined to single out data representing the losses attributed to mites and climate change, allowing us to focus on these critical factors.

3. **Exploratory Focus**:
   - **State and Year Analysis**: By focusing on data that represented colony losses across different states and years, we aimed to explore the geographical and temporal impacts of these factors on bee colonies.

These steps ensured that our dataset was clean, relevant, and well-prepared for subsequent analysis, enabling a focused exploration of the causes of bee colony losses and their relationship with mites and climate change.

@@

# Create the flowchart using DiagrammeR
grViz("
  digraph Data_Analysis_Flowchart {
    graph [layout = dot, rankdir = LR, label = 'Workflow Schematic', fontsize = 75, labelloc = 't']

    node [style = filled, fontsize = 50, width = 6, height = 6, penwidth = 8]

@@

@@

### Analysis of Boxplot

The boxplot illustrates the distribution of values for "LOSS, COLONY COLLAPSE DISORDER" and "LOSS, DEADOUT" across the years 2015 to 2017. A few key observations can be made from the data:

**Central Tendency and Spread**:
   - The median values for both types of losses (represented by the central line in each box) are relatively low across all years, indicating that the majority of losses are concentrated towards the lower end of the scale.

**Yearly Comparison**:
   - For the year 2015, the losses due to Colony Collapse Disorder and Deadout are higher compared to the subsequent years. This can be observed from the height of the boxes and the spread of the data points.
   - There is a noticeable decrease in the losses for both categories in 2016 and 2017. The boxes for these years are smaller and closer to the x-axis, indicating lower values.

**Outliers**:
   - There are significant outliers present in all years for both types of losses. These outliers indicate that while most of the data points are clustered around the lower values, there are instances of very high losses that deviate from the norm.

**Comparison between Loss Types**:
   - The distribution of "LOSS, DEADOUT" appears to have more variability compared to "LOSS, COLONY COLLAPSE DISORDER", particularly in 2015 and 2016. This is indicated by the wider interquartile ranges and more spread out data points.

**Overall Trends**:
   - Overall, the boxplot suggests a decline in both types of losses from 2015 to 2017. However, the presence of outliers in each year indicates that there are still occasional severe loss events.

This highlights the trends and variations in honeybee colony losses due to Colony Collapse Disorder and Deadout over a three-year period, providing insights into the distribution and severity of these losses.

@@

```
### Analysis of Scatterplot

The scatterplot visualizes the distribution of values across different states (represented by State ANSI codes) for the years 2015, 2016, and 2017. Several key observations can be drawn from the data:

**Value Distribution Across States**:
   - The values vary significantly across different states. Most states have relatively low values, but there are notable exceptions with very high values.

**Yearly Trends**:
   - In 2015, there are multiple high values, especially at the lower end of the State ANSI codes. These are represented by the red dots.
   - In 2016, represented by green dots, there are fewer high values, but some extreme outliers are present, such as the one exceeding 200,000.
   - In 2017, indicated by blue dots, the values are generally lower compared to the previous years, with fewer extreme outliers.

**Outliers**:
   - Each year has significant outliers that deviate from the general trend of low values across states. These outliers suggest occasional severe losses in specific states.

**State-Specific Trends**:
   - Certain states (e.g., those with lower ANSI codes) exhibit higher variability and more frequent high values compared to others.
   - States towards the middle and higher end of the ANSI scale generally show more consistent, lower values.

**Comparative Analysis**:
   - The plot indicates that while overall losses may be high in certain states, the distribution and frequency of these high values vary across years. This suggests possible changes in factors affecting bee colony losses over time.

This shows the distribution and variability of bee colony losses across different states over three years, providing insights into state-specific trends and the presence of extreme loss events.

@@

The full analysis is still pending

@@

## References
